---
title: "Introduction to a LSTM-based calibration"
author: "Sima Najafzadehkhoei"
date: today
format: html
vignette: >
  %\VignetteIndexEntry{Introduction to a Lstm-based calibration}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

## Load Required Libraries
We begin by loading the necessary packages. These include the epiworldR suite for simulation, the custom epiworldRcalibrate package for BiLSTM-based parameter inference, and plotting tools.

```{r, results='hide'}
library(tidyverse)
library(epiworldR)
library(epiworldRcalibrate)
library(reticulate)
library(ggplot2)
library(patchwork)
```

```{r}
#| label: _setup-reticulate
#| include: false

# Make sure reticulate doesn't try to create its own Miniconda env on CI
Sys.setenv(RETICULATE_MINICONDA_ENABLED = "FALSE")

# If the workflow exported RETICULATE_PYTHON, point reticulate to it
py <- Sys.getenv("RETICULATE_PYTHON", unset = "")
if (nzchar(py)) {
  try(reticulate::use_python(py, required = FALSE), silent = TRUE)
}

# Check all Python modules you rely on
required_py <- c("torch", "joblib", "numpy")
avail <- vapply(required_py, reticulate::py_module_available, logical(1))
missing_py <- required_py[!avail]
py_ok <- all(avail)

# Keep vignette robust: never hard-stop the whole build
knitr::opts_chunk$set(error = FALSE, warning = TRUE)
```
## Generate a Single Parameter Set
We randomly generate a single set of SIR model parameters to simulate a ground-truth epidemic.

```{r}
set.seed(122)
n_value <- sample(5000:10000, 1)
preval <- runif(1, 0.007, 0.02)
crate <- runif(1, 1, 5)
recov <- runif(1, 0.071, 0.25)
R0 <- runif(1, 1.1, 5)
ptran <- (R0 * recov / crate)

true_params <- data.frame(
  n = n_value,
  prevalence = preval,
  contact_rate = crate,
  transmission_rate = ptran,
  recovery_rate = recov,
  R0 = R0
)
print(true_params)
```

## Run True Simulation
We now simulate the epidemic using the SIR model with our true parameters.

```{r}
ndays <- 60
true_model <- ModelSIRCONN(
  name = "true_simulation",
  n = true_params$n,
  prevalence = true_params$prevalence,
  contact_rate = true_params$contact_rate,
  transmission_rate = true_params$transmission_rate,
  recovery_rate = true_params$recovery_rate
)
run(true_model, ndays = ndays)
incidence_data <- plot_incidence(true_model, plot = TRUE)
incidence_ts <- incidence_data[, 1]
print(incidence_ts)
```

## Quick Parameter Estimation with BiLSTM
We use the convenient one-step function to estimate SIR parameters from the simulated incidence data. This automatically initializes the BiLSTM model and makes predictions.

```{r}
# Quick estimation using the one-step function
# This automatically loads the pre-trained BiLSTM model and estimates parameters
lstm_predictions <- quick_estimate_sir(
  daily_cases = incidence_ts,
  population_size = true_params$n,
  recovery_rate = true_params$recovery_rate
)

cat("LSTM Parameter Predictions:\n")
print(lstm_predictions)

# Extract individual parameters for downstream analysis
lstm_params <- data.frame(
  n = true_params$n,
  prevalence = true_params$prevalence,
  contact_rate = lstm_predictions[["crate"]],
  transmission_rate = lstm_predictions[["ptran"]],
  recovery_rate = true_params$recovery_rate,
  R0 = lstm_predictions[["R0"]]
)
print(lstm_params)
```

## Understanding Data Preprocessing
Let's examine how the raw incidence data is preprocessed before being fed to the neural network.

```{r}
# Show how the incidence data is transformed (first 10 days)
preprocessing_demo <- show_preprocessing(incidence_ts[1:10])
print(preprocessing_demo)

cat("\nThe model converts raw case counts to percentage changes:\n")
cat("- Day 0: Always 0 (reference point)\n")
cat("- Day i: (count[i] - count[i-1]) / count[i-1] if count[i-1] > 0\n")
cat("- Day i: 0 if count[i-1] == 0 (handles division by zero)\n")
```

## Compare True vs LSTM Parameters
This block prepares a table of both true and predicted parameters for further comparison and simulation.

```{r}
params_comparison <- bind_rows(
  true_params %>% mutate(param_type = "true"),
  lstm_params %>% mutate(param_type = "lstm")
) %>% mutate(sim_id = 1, run_id = row_number())
print(params_comparison)
```

## Run Multiple Simulations
We run multiple replicates of the SIR model for each parameter set to understand variability and average behavior.

```{r}
n_reps <- 100
all_simulation_results <- tibble()

for (i in seq_len(nrow(params_comparison))) {
  row <- params_comparison[i, ]
  forward_model <- ModelSIRCONN(
    name = paste0("forward_", row$param_type),
    n = row$n,
    prevalence = row$prevalence,
    contact_rate = row$contact_rate,
    transmission_rate = row$transmission_rate,
    recovery_rate = row$recovery_rate
  )
  saver <- make_saver("total_hist")
  run_multiple(forward_model, ndays = ndays, nsims = n_reps, saver = saver, nthreads = 8)
  results <- run_multiple_get_results(forward_model, nthreads = 8)
  sim_data <- results$total_hist %>%
    group_by(date, state) %>%
    summarize(mean_count = mean(counts),
              ci_lower = quantile(counts, 0.025),
              ci_upper = quantile(counts, 0.975), .groups = "drop") %>%
    mutate(param_type = row$param_type, sim_id = row$sim_id)
  all_simulation_results <- bind_rows(all_simulation_results, sim_data)
}
```

## Visualize Results
We generate comparative plots of Susceptible, Infected, and Recovered individuals over time, including confidence intervals for the LSTM-based predictions.

```{r}
method_colors <- c("true" = "#440154FF", "lstm" = "#35B779FF")

create_sir_plot <- function(data, state_name, title) {
  plot_data <- data %>% filter(state == state_name)
  ggplot(plot_data, aes(x = date, color = param_type)) +
    geom_ribbon(data = plot_data %>% filter(param_type == "lstm"),
                aes(ymin = ci_lower, ymax = ci_upper, fill = param_type), alpha = 0.2, color = NA) +
    geom_line(aes(y = mean_count), linewidth = 1.2) +
    scale_color_manual(values = method_colors) +
    scale_fill_manual(values = method_colors) +
    labs(title = title, x = "Day", y = "Count (95% CI)", color = "Method", fill = "Method") +
    theme_minimal() +
    theme(legend.position = "bottom", plot.title = element_text(size = 12, hjust = 0.5))
}

p_susceptible <- create_sir_plot(all_simulation_results, "Susceptible", "Susceptible Over Time")
p_infected <- create_sir_plot(all_simulation_results, "Infected", "Infected Over Time")
p_recovered <- create_sir_plot(all_simulation_results, "Recovered", "Recovered Over Time")

(p_susceptible / p_infected / p_recovered) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "SIR Model Comparison: True vs LSTM Parameters",
                  subtitle = paste0("Single parameter set with ", n_reps, " replicates each"))
```

## Calculate Bias
We compare the inferred parameters to their ground truth values and summarize differences in epidemic trajectories.

### Parameter Bias Table (Static Parameter Comparison)

This table compares the **true parameter values** (used to generate the epidemic data) with the **estimated values** predicted by the LSTM model. It includes:

* **Bias**: the raw difference `LSTM - True`
* **Relative Bias (%)**: the percent difference relative to the true value

```{r}
# Construct parameter comparison table
param_bias <- data.frame(
  Parameter = c("Contact Rate", "Transmission Rate", "R0"),
  True = c(true_params$contact_rate, true_params$transmission_rate, true_params$R0),
  LSTM = c(lstm_params$contact_rate, lstm_params$transmission_rate, lstm_params$R0)
) %>%
  mutate(
    Bias = LSTM - True,
    Relative_Bias = (LSTM - True) / True * 100
  )

# Show the table
print(param_bias)
```

**Interpretation**:

* A **Bias** close to 0 means the LSTM prediction is very close to the true value.
* A **Relative Bias** close to 0% indicates high accuracy.
* Negative bias means underestimation, and positive means overestimation.

### Time Series Bias (Infected Trajectory Comparison)

This part compares the **predicted epidemic curves** (Infected over time) generated using the true parameters and those estimated by LSTM.

```{r}
# Compare infected time series from both simulations
trajectory_comparison <- all_simulation_results %>%
  filter(state == "Infected") %>%
  select(date, param_type, mean_count) %>%
  pivot_wider(names_from = param_type, values_from = mean_count) %>%
  mutate(
    bias = lstm - true,
    relative_bias = (lstm - true) / true * 100
  )

# View trajectory comparison for each time step
print(trajectory_comparison)
```

### Summary Statistics for Infected Curve

This summarizes the prediction performance across all days:

```{r}
# Summarize the overall bias metrics
bias_summary <- trajectory_comparison %>%
  summarize(
    mean_bias = mean(bias, na.rm = TRUE),                  # average error
    mean_abs_bias = mean(abs(bias), na.rm = TRUE)          # average magnitude of error
  )

print(bias_summary)
```

## Advanced: Manual Model Management
For users who need more control over model loading and memory management:

```{r}
# Check model status (quick_estimate_sir automatically loaded it)
status <- check_model_status()
cat("Model loaded:", status$loaded, "\n")
cat("Model directory:", status$model_directory, "\n")

# Alternative: Manual initialization and estimation
# This gives the same results but with more explicit control
if (FALSE) {  # Set to TRUE if you want to run this section
  
  # Step 1: Initialize manually
  init_bilstm_model()
  
  # Step 2: Estimate manually  
  manual_predictions <- estimate_sir_parameters(
    daily_cases = incidence_ts,
    population_size = true_params$n,
    recovery_rate = true_params$recovery_rate
  )
  
  # Verify they match quick_estimate_sir results
  cat("Manual vs Quick estimation comparison:\n")
  comparison <- data.frame(
    Parameter = c("ptran", "crate", "R0"),
    quick_estimate = c(lstm_predictions[["ptran"]], lstm_predictions[["crate"]], lstm_predictions[["R0"]]),
    manual_estimate = c(manual_predictions[["ptran"]], manual_predictions[["crate"]], manual_predictions[["R0"]])
  )
  print(comparison)
}

# Optional: Clean up model from memory to free resources
# cleanup_model()
```

## Save Results
All analysis results, including parameter estimates, simulation outputs, and bias summaries are saved for future use.

```{r}
results_list <- list(
  true_parameters = true_params,
  lstm_parameters = lstm_params,
  lstm_predictions = lstm_predictions,
  parameter_comparison = params_comparison,
  simulation_results = all_simulation_results,
  parameter_bias = param_bias,
  trajectory_bias = bias_summary,
  preprocessing_demo = preprocessing_demo
)

str(results_list, 1)
```

## Key Function Summary

The `epiworldRcalibrate` package provides these main functions:

* **`quick_estimate_sir()`**: One-step function that automatically initializes and estimates (recommended for most users)
* **`init_bilstm_model()`**: Initialize the pre-trained BiLSTM model manually
* **`estimate_sir_parameters()`**: Estimate SIR parameters from 61 days of incidence data (requires manual initialization)
* **`show_preprocessing()`**: Demonstrate how raw data is transformed for the model
* **`check_model_status()`**: Check if model is loaded and get directory info
* **`cleanup_model()`**: Remove model from memory to free resources

## Quick Start Workflow

For most users, the simplest workflow is:

```{r, eval=FALSE}
# 1. Load your 61-day incidence data
my_incidence <- c(...)  # exactly 61 consecutive daily case counts

# 2. Set your population and recovery rate
population_size <- 10000
recovery_rate <- 1/7  # 7-day infectious period

# 3. Get parameter estimates in one step
params <- quick_estimate_sir(my_incidence, population_size, recovery_rate)

# 4. Use the estimated parameters in your SIR model
print(params)  # ptran, crate, R0
```

## Conclusion
This vignette illustrates how to simulate a ground-truth epidemic using `epiworldR`, apply a BiLSTM-based parameter inference method with `epiworldRcalibrate`, and compare predicted versus true epidemic dynamics. The `quick_estimate_sir()` function provides the most convenient entry point for parameter estimation, automatically handling model initialization and providing robust parameter estimates for SIR model calibration.
