---
title: "LSTM-Calibrated SIR Model for COVID-19 Dynamics"
subtitle: "Utah COVID-19 Case Prediction with Uncertainty Quantification"
author: "Sima Najafzadehkhoei"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
    embed-resources: true
editor: visual
---

## Introduction

This analysis demonstrates an innovative approach to epidemic modeling by combining deep learning with traditional compartmental models. We use a BiLSTM (Bidirectional Long Short-Term Memory) neural network to calibrate parameters for a stochastic SIR (Susceptible-Infected-Recovered) model, then run multiple simulations to quantify prediction uncertainty.

### Workflow Overview

1.  Download recent COVID-19 case data from Utah
2.  Use BiLSTM to estimate SIR model parameters from 61-day incidence patterns
3.  Run 2,000 stochastic SIR simulations with calibrated parameters
4.  Extract daily infection counts and compute 95% confidence intervals
5.  Compare model predictions against observed data

## Setup

Load required packages for data manipulation, modeling, and visualization:

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(epiworldR)
  library(epiworldRcalibrate)
})
```

## Data Acquisition

### Download Functions

We create helper functions to download and process Utah's COVID-19 surveillance data:

```{r data-helpers}

data("utah_covid_data")

# Now you can use it directly
head(utah_covid_data)
summary(utah_covid_data)
```

### Load Recent Case Data

Retrieve the most recent 61 days of daily case counts:

```{r load-data}
get_covid_data <- function(n_days) {
  
  # Use packaged data
  covid_data <- utah_covid_data
  
  # Filter to last n_days
  last_date <- max(covid_data$Date, na.rm = TRUE)
  covid_data <- subset(covid_data, Date > (last_date - n_days)) %>% 
    dplyr::arrange(Date)
  
  stopifnot("Daily.Cases" %in% names(covid_data))
  covid_data
}
n_days <- 61
covid_data <- get_covid_data(n_days)
incidence <- covid_data$Daily.Cases
incidence_vec <- incidence

stopifnot(length(incidence) == n_days)

cat("Date range:", as.character(min(covid_data$Date)), "to", 
    as.character(max(covid_data$Date)), "\n")
cat("Total cases observed:", sum(incidence), "\n")
cat("Mean daily cases:", round(mean(incidence), 1), "\n")
```

The BiLSTM model requires exactly 61 consecutive days of incidence data. This window length captures approximately two months of epidemic dynamics, sufficient to identify transmission patterns while remaining computationally efficient.

## BiLSTM Parameter Calibration

### Model Parameters

The SIR model requires three key parameters:

-   **Transmission probability (ptran)**: Probability of infection per contact
-   **Contact rate (crate)**: Average number of contacts per person per day\
-   **Basic reproduction number (R₀)**: Expected secondary infections per case

### Running Calibration

We define the population size and recovery rate, then use the BiLSTM to estimate parameters:

```{r calibrate}
# Model configuration
N     <- 0000  # Population size
recov <- 1 / 7     # Recovery rate (7-day infectious period)

# LSTM calibration
lstm_predictions <- calibrate_sir(
  daily_cases = incidence_vec,
  population_size = N,
  recovery_rate = recov
)

print(lstm_predictions)
```

The BiLSTM neural network has learned relationships between incidence time series patterns and underlying transmission dynamics from training data. It processes the 61-day sequence bidirectionally to capture both past and future context when estimating parameters.

### Parameter Consistency

The contact rate is recalculated using the fundamental epidemiological relationship:

$$\text{contact rate} = \frac{R_0 \times \text{recovery rate}}{\text{transmission probability}}$$

This ensures internal consistency between the three parameters.

### Initial Conditions

```{r initial-conditions}
init_infected <- incidence[1]
prev <- init_infected / N

cat("Initial infected count:", init_infected, "\n")
cat("Initial prevalence:", round(prev * 100, 4), "%\n")
```

## Stochastic SIR Simulation

### Building the Model

We construct a contact-network SIR model using the calibrated parameters:

```{r build-model}
model <- ModelSIRCONN(
  name              = "Utah LSTM-calibrated SIR",
  n                 = N,
  prevalence        = prev,
  contact_rate      = lstm_predictions[["crate"]],
  transmission_rate = lstm_predictions[["ptran"]],
  recovery_rate     = recov
)
```

The `ModelSIRCONN` structure simulates disease transmission through a contact network, providing more realistic dynamics than simple mass-action models.

### Multiple Simulation Runs

To quantify uncertainty, we run 2,000 independent stochastic realizations:

```{r run-simulations}
saver <- make_saver("transition")

run_multiple(
  model,
  ndays    = n_days - 1,   # 60 steps yields 61 time points (days 0-60)
  nsims    = 2000,
  saver    = saver,
  nthreads = 12
)

sim_results <- run_multiple_get_results(
  model,
  nthreads = 12,
  freader = data.table::fread
)
```

Each simulation represents one possible trajectory of the epidemic. By running many simulations, we capture the stochastic variability inherent in transmission processes.

## Extracting Daily Infections

### Transition Matrices

The model tracks all state transitions. We focus on Susceptible → Infected transitions, which represent new daily infections:

```{r extract-transitions}
cat("=== Extracting S→I transitions ===\n")

transitions <- sim_results$transition %>%
  filter(from == "Susceptible", to == "Infected") %>%
  arrange(sim_num, date)

cat("Dimensions:", nrow(transitions), "rows ×", ncol(transitions), "columns\n")
head(transitions, 10)
```

### Computing Confidence Intervals

For each day, we calculate quantiles across all 2,000 simulations:

```{r calculate-quantiles}
cat("=== Calculating 95% CI (2.5th and 97.5th percentiles) ===\n")

quantiles_df <- transitions %>%
  group_by(date) %>%
  summarize(
    lower_ci = quantile(counts, 0.025),
    upper_ci = quantile(counts, 0.975),
    median   = quantile(counts, 0.5),
    mean     = mean(counts),
    .groups  = "drop"
  )

head(quantiles_df, 10)
```

The 95% confidence interval represents the range containing 95% of simulation outcomes, providing a measure of prediction uncertainty.

## Visualization

### Merge with Observed Data

```{r prepare-plot-data}
plot_df <- quantiles_df %>%
  mutate(
    Date           = covid_data$Date,
    observed_cases = incidence
  )

head(plot_df)
```

### Plot Results with Confidence Intervals

```{r plot-results, fig.width=11, fig.height=7}
p_with_ci <- ggplot(plot_df, aes(x = Date)) +
  # 95% confidence interval
  geom_ribbon(
    aes(ymin = lower_ci, ymax = upper_ci),
    fill = "red",
    alpha = 0.3
  ) +
  # Model median
  geom_line(
    aes(y = median, color = "Model median (S→I)"),
    linewidth = 1.2
  ) +
  # Observed cases
  geom_line(
    aes(y = observed_cases, color = "Observed cases"),
    linewidth = 1.4
  ) +
  geom_point(
    aes(y = observed_cases, color = "Observed cases"),
    size = 2
  ) +
  scale_color_manual(
    values = c(
      "Model median (S→I)" = "red",
      "Observed cases"     = "blue"
    )
  ) +
  labs(
    title = "Daily Infected: Observed vs Model with 95% CI",
    subtitle = paste0(
      "Population: ", format(N, big.mark = ","),
      " | Contact rate: ", round(lstm_predictions[["crate"]], 3),
      " | Trans. prob: ", round(lstm_predictions[["ptran"]], 3),
      " | R₀: ", round(lstm_predictions[["R0"]], 2)
    ),
    x = "Date",
    y = "Daily Infection Counts",
    color = "",
    fill = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(p_with_ci)
```

The red shaded region represents the 95% confidence interval from 2,000 simulations. The red line shows the median prediction, while blue points are observed case counts.

## Model Performance

### Goodness-of-Fit Metrics

```{r performance-metrics}
# CI coverage
coverage <- plot_df %>%
  mutate(in_ci = observed_cases >= lower_ci & observed_cases <= upper_ci) %>%
  summarize(
    coverage_rate = mean(in_ci) * 100,
    n_days = n(),
    days_in_ci = sum(in_ci)
  )

# Error metrics
rmse <- sqrt(mean((plot_df$median - plot_df$observed_cases)^2))
mae <- mean(abs(plot_df$median - plot_df$observed_cases))
mape <- mean(abs((plot_df$observed_cases - plot_df$median) / 
                   plot_df$observed_cases)) * 100

cat("=== Model Performance Metrics ===\n\n")
cat("95% CI Coverage:", round(coverage$coverage_rate, 1), "%\n")
cat("  (", coverage$days_in_ci, "out of", coverage$n_days, "days within CI)\n\n")
cat("Root Mean Squared Error (RMSE):", round(rmse, 2), "cases/day\n")
cat("Mean Absolute Error (MAE):", round(mae, 2), "cases/day\n")
cat("Mean Absolute Percentage Error (MAPE):", round(mape, 1), "%\n")
```

### Interpretation

**Coverage**: A well-calibrated model should have approximately 95% of observed values within the 95% confidence interval. Lower coverage suggests underestimated uncertainty; higher coverage suggests overestimated uncertainty.

**RMSE**: Measures average prediction error magnitude, giving more weight to large errors.

**MAE**: Measures average absolute error without emphasizing outliers.

**MAPE**: Percentage error metric that is scale-independent.

## Discussion

### Key Results

The BiLSTM-calibrated SIR model successfully captured COVID-19 transmission dynamics in Utah:

-   Estimated transmission probability: **`r round(lstm_predictions[["ptran"]], 3)`**
-   Estimated contact rate: **`r round(lstm_predictions[["crate"]], 3)`** contacts/person/day
-   Basic reproduction number: **R₀ = `r round(lstm_predictions[["R0"]], 2)`**

The model achieved **`r round(coverage$coverage_rate, 1)`%** coverage with a 95% confidence interval and mean absolute error of **`r round(mae, 2)`** cases per day.

### Advantages of This Approach

1.  **Hybrid methodology**: Combines neural network pattern recognition with mechanistic epidemiological modeling
2.  **Interpretable parameters**: Outputs meaningful epidemiological quantities (R₀, contact rates)
3.  **Uncertainty quantification**: Stochastic simulations provide prediction intervals
4.  **Real-time capability**: Can be updated as new surveillance data becomes available
5.  **Minimal assumptions**: BiLSTM learns patterns from data rather than requiring detailed knowledge of transmission mechanisms

### Limitations

-   **Homogeneous mixing**: Assumes uniform contact patterns across the population
-   **Constant parameters**: Does not capture time-varying transmission due to interventions or behavioral changes
-   **Closed population**: No births, deaths, or migration
-   **Fixed recovery period**: Assumes all individuals recover at the same rate
-   **No spatial structure**: Treats the entire state as a single well-mixed population

### Future Directions

-   Incorporate time-varying parameters to model interventions (lockdowns, mask mandates)
-   Extend to SEIR model including exposed (latent) compartment

## Conclusion

This analysis demonstrates that BiLSTM neural networks can effectively calibrate traditional compartmental epidemic models from observed incidence data. The resulting stochastic SIR simulations provide both point predictions and uncertainty quantification, offering a powerful tool for public health decision-making.

The approach is particularly valuable when detailed knowledge of transmission mechanisms is limited but historical incidence data is available. By learning patterns from data, the BiLSTM can estimate parameters that would otherwise require extensive epidemiological investigation.
