---
title: "Introduction to a Lstm-based calibration"
author: "Sima Najafzadehkhoei"
date: today
format: html
vignette: >
  %\VignetteIndexEntry{Introduction to a Lstm-based calibration}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

## Load Required Libraries

We use `epiworldR` for simulation, `epiworldRcalibrate` for BiLSTM-based parameter inference, and tidy tools for data handling and plotting.

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(epiworldR)
  library(epiworldRcalibrate)
  library(ggplot2)
  library(patchwork)
  library(zoo)
  library(reticulate)
})
```
Fetch Utah DHHS COVID-19 Trends data
We pull the Trends_Epidemic CSV from the official Utah ZIP, then view the data to confirm columns.

```{r}
get_data_from_url <- function(data_url, target_file) {
  tryCatch({
    temp_file <- tempfile()
    download.file(data_url, temp_file, mode = "wb")
    datafile_name <- grep(
      target_file,
      unzip(temp_file, list = TRUE)$Name,
      ignore.case = TRUE,
      value = TRUE
    )
    stopifnot(length(datafile_name) >= 1)
    datafile <- unz(temp_file, datafile_name[1])
    collected_data <- read.csv(datafile)
    return(collected_data)
  },
  error = function(cond) {
    message("Error: ", conditionMessage(cond)); return(-1)
  },
  warning = function(cond) {
    message("Warning: ", conditionMessage(cond)); return(-2)
  },
  finally = { unlink(temp_file) })
}

get_covid_data <- function(n_days = 120) {
  data_url <- "https://coronavirus-dashboard.utah.gov/Utah_COVID19_data.zip"
  target_file_regex <- "Trends_Epidemic+"
  covid_data <- get_data_from_url(data_url, target_file_regex)
  stopifnot(is.data.frame(covid_data))
  covid_data$Date <- as.Date(covid_data$Date)
  last_date <- max(covid_data$Date, na.rm = TRUE)
  covid_data[covid_data$Date > (last_date - n_days), ]
}

raw <- get_covid_data(120)
head(raw, 10)
tail(raw, 10)
str(raw)
```
Quick sanity plots
```{r}
p1 <- ggplot(raw, aes(Date, Daily.Cases)) +
  geom_col(width = 1) +
  labs(title = "Utah Daily Cases (last 120 days)", x = NULL, y = "Daily cases") +
  theme_minimal()

p2 <- ggplot(raw, aes(Date, Smoothed.3.Day.Moving.Average)) +
  geom_line() +
  labs(title = "Smoothed 3-day Moving Average", x = NULL, y = "Smoothed cases") +
  theme_minimal()

p1 / p2
```
Build a smooth 61-day incidence window
We use the smoothed series (already plateau/stable in your excerpt). If needed, you can swap in a 7-day moving average of Daily.Cases.

```{r}
# Ensure chronological order and compute optional 7-day MA
raw <- raw %>%
  arrange(Date) %>%
  mutate(ma7 = zoo::rollmean(Daily.Cases, 7, fill = NA, align = "right"))

# Pick the last 61 days with non-NA smoothed values
seg <- raw %>% filter(!is.na(Smoothed.3.Day.Moving.Average)) %>% tail(61)

# Visual check of the chosen window
ggplot(seg, aes(Date, Daily.Cases)) +
  geom_col(width = 1, alpha = 0.6) +
  geom_line(aes(y = Smoothed.3.Day.Moving.Average)) +
  labs(title = "Chosen 61-day window (Utah DHHS Trends)", x = NULL, y = "Daily cases / smoothed") +
  theme_minimal()

# Incidence vector for calibration (use smoothed values)
incidence <- as.numeric(raw$Daily.Cases)[0:61]
stopifnot(length(incidence) == 61L)
```
Predict SIR parameters using the BiLSTM
Initialize your pretrained model and infer parameters from the 61-day incidence.

```{r}
# Adjust path to your installed model directory as documented in README
# Example: init_bilstm_model("~/Desktop/epiworldRcalibrate_fixed/epiworldRcalibrate/inst/models")
init_bilstm_model(system.file("models", package = "epiworldRcalibrate", mustWork = FALSE))

N     <- 100000        # ABM population scale (covariate)
recov <- 1/7           # ~7-day infectious period

lstm_predictions <- predict_sir_bilstm(incidence, n = N, recov = recov)
lstm_predictions
```
# $ptran, $crate, $R0
For completeness, here’s a tidy params row we’ll pass to the simulator:

```{r}
lstm_params <- tibble::tibble(
  n                 = N,
  prevalence        = NA_real_,           # will be set from last observed cases
  contact_rate      = lstm_predictions[["crate"]],
  transmission_rate = lstm_predictions[["ptran"]],
  recovery_rate     = recov,
  R0                = lstm_predictions[["R0"]]
)
lstm_params
```
14-day forecast with epiworldR
We seed the SIR model using an inferred initial prevalence from the last observed day. This requires an ascertainment assumption to translate reported cases to infections.

```{r}
ascertainment <- 0.25  # assumed fraction of infections that are reported cases
last_daily    <- tail(incidence, 1)
init_infected <- max(1, round(last_daily / ascertainment))
prev          <- init_infected / N

forecast_model <- ModelSIRCONN(
  name              = "COVID-19 (Utah forecast)",
  n                 = N,
  prevalence        = prev,
  contact_rate      = lstm_params$contact_rate,
  transmission_rate = lstm_params$transmission_rate,
  recovery_rate     = lstm_params$recovery_rate
)

horizon <- 14L
n_reps  <- 200L
saver   <- make_saver("total_hist")

set.seed(123)
run_multiple(forecast_model, ndays = horizon, nsims = n_reps, saver = saver, nthreads = 8)
results <- run_multiple_get_results(forecast_model, nthreads = 8)

# Aggregate across simulations to get uncertainty bands
fc <- results$total_hist %>%
  dplyr::filter(state == "Infected") %>%
  group_by(date) %>%
  summarise(
    mean = mean(counts),
    p50  = median(counts),
    lo50 = quantile(counts, 0.25),
    hi50 = quantile(counts, 0.75),
    lo95 = quantile(counts, 0.025),
    hi95 = quantile(counts, 0.975),
    .groups = "drop"
  ) %>%
  mutate(calendar_date = max(seg$Date) + date)
```
Visualize observed vs forecast
Observed bars = reported daily cases (61-day window). Lines/bands = simulated infected over the next 14 days under the LSTM-calibrated SIR.

```{r}
p_obs <- ggplot(seg, aes(Date, Daily.Cases)) +
  geom_col(width = 1, alpha = 0.6) +
  labs(x = NULL, y = "Daily cases (observed)") +
  theme_minimal() +
  ggtitle("Observed (Utah DHHS)")

p_fc <- ggplot() +
  geom_ribbon(data = fc, aes(calendar_date, ymin = lo95, ymax = hi95), alpha = 0.15) +
  geom_ribbon(data = fc, aes(calendar_date, ymin = lo50, ymax = hi50), alpha = 0.25) +
  geom_line(data = fc, aes(calendar_date, p50), linewidth = 1) +
  labs(x = NULL, y = "Simulated infected (forecast)") +
  theme_minimal() +
  ggtitle("14-day SIR forecast (LSTM-calibrated)")

p_obs / p_fc + plot_annotation(
  title = "Utah COVID-19: 14-day forecast via epiworldR + BiLSTM calibration",
  subtitle = "Incidence window: last 61 smoothed days from Trends_Epidemic"
)
```
Save key outputs
```{r}
bias_summary <- tibble::tibble() # placeholder (no ground-truth future yet)

results_list <- list(
  lstm_parameters = lstm_params,
  incidence_window = seg %>% select(Date, Daily.Cases, Smoothed.3.Day.Moving.Average),
  forecast_summary = fc,
  assumptions = list(N = N, recov = recov, ascertainment = ascertainment),
  diagnostics = list(bias_future = bias_summary)
)

str(results_list, 1)
```
Notes
Data source: Utah DHHS Trends_Epidemic inside Utah_COVID19_data.zip.

“No intervention” filter: we used Smoothed.3.Day.Moving.Average and selected the last 61 non-NA days; this is typically smooth (plateau) and stable.

Ascertainment: tune ascertainment if you have a better estimate; it affects initial prevalence and thus forecast magnitude.

Model path: ensure init_bilstm_model() points to your package’s inst/models (or wherever your .pt and scaler .pkls live).
